Practical Optional Types for Clojure

### 2015-02-26 formal-model.tex
explain T-Do better

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals and let expressions.
All binding forms introduce fresh variables.
Values include booleans, \nil{}, class literals, keywords, 
constants and closures. \false{} and \nil{} are logically false, otherwise values are true.

@@

@@

We extend our model with sequencing expressions and errors, where {\errorvalv{}}
models Clojure's throw special form.

@@

Now we can check example~\ref{example:doexception}.
We check the test with T-App.
$$
\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(number? x)}}{\clj{B}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
$$
By T-Nil and subsumption we derive the same information in both propositions.
$$
\judgement{\isprop{\clj{Num}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{Num}}{\x{}}}{\notprop{\clj{Num}}{\x{}}}}{\emptyobject{}}
$$
By T-Error we can assume anything about the return of an error.
$$
\judgement{\notprop{\clj{Num}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
$$
Using the above as premises to T-If we learn {\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}
in both propositions.
$$
\judgement{\isprop{\clj{(U nil Num)}}{\x{}}}{\hastype{\clj{(if (number? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
          {\filterset{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{Num}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
$$
By T-Local we know \clj{x} is a \clj{Num}.
$$
\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\x{}}
$$
\clj{inc} requires a \clj{Num} argument which we just proved.
$$
\judgement{\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{Num}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\emptymap{}}
    &\mbox{Values} \\
    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
    &\mbox{Types} \\
    \auxhmapsyntax{}\\
    \pesyntax{}   &::=& \ldots \alt {\keype{\k{}}}
                  &\mbox{Path Elements}
  \end{altgrammar}
  $$
  \caption{HMap Terms and Types}
  \label{main:figure:hmapsyntax}
\end{figure}

\begin{figure}
  \footnotesize
  \begin{mathpar}
    {\TGetHMap}

@@

    {\TAssoc}
  \end{mathpar}
  \caption{HMap Typing Rules}
  \label{main:figure:hmaptyping}
\end{figure}

@@

Figure~\ref{main:figure:hmaptyping} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
T-Assoc extends a HMap with a mandatory entry.
Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
and \assocliteral{}.

>>>>

Syntax is given in figure~\ref{main:figure:termsyntax}. Expressions include variables, values,
application, abstractions, conditionals and let expressions.
All binding forms introduce fresh variables.
Values include booleans, \nil{}, class literals, keywords, 
constants and closures. 
Value environments map local bindings to values.
Closures contain the value environment \openv{} when the function was created,
which is used during beta reduction to ensure the abstraction respects lexical scoping.
\false{} and \nil{} are logically false, otherwise values are true.

@@

For example,
B-IfTrue evaluates \e{1} in the environment \openv{} with the result \v{1}.
If \v{1} is not \nil{} or \false{} then the conditional returns the result of evaluating the then branch \e{2}
in \openv{}.

@@

We extend our model with sequencing expressions and errors, where {\errorvalv{}}
models the result of calling Clojure's \clj{throw} special form
with some \clj{Throwable}.

@@

Now we can check example~\ref{example:doexception}.
We check the test with T-App.
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(n? x)}}{\clj{B}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
$$
By T-Nil and subsumption we derive the same information in both propositions.
$$
\judgement{\isprop{\clj{N}}{\x{}}}{\hastype{\clj{nil}}{\clj{nil}}}{\filterset{\isprop{\clj{N}}{\x{}}}{\notprop{\clj{N}}{\x{}}}}{\emptyobject{}}
$$
By T-Error we can assume anything about the return of an error.
$$
\judgement{\notprop{\clj{N}}{\x{}}}{\hastype{\errorvalv{}}{\Bot}}{\filterset{\botprop{}}{\botprop{}}}{\emptyobject{}}
$$
Using the above as premises to T-If we learn {\orprop{\isprop{\clj{Num}}{\x{}}}{\botprop{}}}
in both propositions.
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(if (n? x) nil}\ {\errorvalv{}{\clj{)}}}}{\clj{B}}}
          {\filterset{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\orprop{\notprop{\clj{N}}{\x{}}}{\botprop{}}}}{\emptyobject{}}
$$
By T-Local we know \clj{x} is a \clj{N}.
$$
\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{x}}{\clj{N}}}{\filterset{\notprop{\clj{(U nil false)}}{\x{}}}{\isprop{\clj{(U nil false)}}{\x{}}}}{\x{}}
$$
By T-App,
\clj{inc} requires a \clj{Num} argument which we just proved.
$$
\judgement{\orprop{\isprop{\clj{N}}{\x{}}}{\botprop{}}}{\hastype{\clj{(inc x)}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$
Finally we conclude with T-Do using the conditional and increment expressions as premises.
$$
\judgement{\isprop{\clj{(U nil N)}}{\x{}}}{\hastype{\clj{(do (if (n? x) nil}\ {\errorvalv{}}{\clj{) (inc x))}}}{\clj{N}}}{\filterset{\topprop{}}{\topprop{}}}{\emptyobject{}}
$$

@@

\begin{figure}
  \footnotesize
  $$
  \begin{altgrammar}
    \e{} &::=& \ldots \alt \hmapexpressionsyntax{}
    &\mbox{Expressions} \\
    \v{} &::=& \ldots \alt {\emptymap{}}
    &\mbox{Values} \\
    \t{} &::=& \ldots \alt {\HMapgeneric {\mandatory{}} {\absent{}}}
    &\mbox{Types} \\
    \auxhmapsyntax{}\\
    \pesyntax{}   &::=& \ldots \alt {\keype{\k{}}}
                  &\mbox{Path Elements}
  \end{altgrammar}
  $$
  \begin{mathpar}
    {\TGetHMap}

@@

    {\TAssoc}
  \end{mathpar}
  \caption{HMap Terms, Types and Typing Rules}
  \label{main:figure:hmapsyntax}
\end{figure}

@@

Figure~\ref{main:figure:hmapsyntax} contains the typing rules. T-GetHMap models a lookup
that will certainly succeed, T-GetHMapAbsent a lookup that will certainly fail
and T-GetHMapPartialDefault a lookup with unknown results.
T-Assoc extends a HMap with a mandatory entry.
Figure~\ref{main:figure:hmapsem} contains the semantics for \getliteral{}
and \assocliteral{}.
